Lonely and online: an investigation of social media use and social isolation in China

### 2023-10-08 Social Isolation in China.qmd
prepared for t and f submission

---
title: "The Spatial Relationship between Social Media, Digital Connectedness, and Social Isolation in China: Revisting Cyber-Optimism"

date: '`r format(Sys.time(), "%B %d, %Y")`'
abstract: "Early research and popular discourse on the promise of the internet's impact on social isolation was optimistic. Since these early predictions, research has become less positive, usually pointing to the negative role of social media. However, little is known about this relationship in highly controlled, autocratic social media environments. Using data from a nationwide survey of Chinese citizens, we do find that the more citizens digitally connect with others, the less personally isolated they feel. Conversely, though, the more likely people are to move relationships from online to offline, and the more they use social media, the more personally isolated they feel. We also find that this relationship interacts with urbanicity - being in rural areas tends to amplify the effect. Overall, our findings suggest that institutional context may not play an important role in the social media - social isolation relationship."
keywords: 
  - Social Isolation
  - Social Media
  - Digital Communication
  - China
format: pdf
execute: 
  echo: false
  warning: false
keep-tex: true
bibliography: references.bib
csl: sage-harvard.csl
---

@@

Research is quite mixed when it comes to identifying the relationship between social media use and feelings of isolationism. Early research suggested that the internet would build social capital, or an interconnectedness between people, and that social media platforms were the ultimate way to bring people together. Since then, research indicates that the effect may be the opposite among at least some part of the population. Social media may encourage social isolation, and ultimately, social media users may feel more disconnected from others than those who do not use social media. We know very little about this phenomenon in China. China presents an interesting case because the institutional structures, level of development and the cultural context are all quite different than the areas studied in most existing studies. This allows us to test whether such features moderate or exacerbate the relationship between social media use and social isolation. To fully test this theory, we employ a large survey of Chinese internet users and test a variety of regression specifications.[^1] The main result is that social media use is robustly negatively related to social isolation, though the nature of this relationship changes depending on the urbanicity of the respondent - being in a rural area amplifies this relationship.

@@

The existing literature on social isolation and social media use generally finds a negative relationship between the two. However, there are some remaining questions regarding whether this relationship remains significant in different cultural and regulatory environments. Based on the existing literature, we then develop some testable hypotheses regarding how the relationship between usage and isolation may vary in the context of China. The result of these hypotheses tests will help to better understand the mechanism that links the two variables.

@@

The most important of these contextual factor in China is that social media space that is highly controlled online environment. Many people are aware that the Chinese government heavily censors politically sensitive content on online platforms. However, fewer are aware that the Chinese government spends as much, if not more, effort to clean up non-political as negative or distressing content on the Chinese internet, such as scams, instances of cyber bullying, disturbing videos, and vulgar content [@wang2020]. This control impulse also extends content the government views as spiritually polluting, such as pornography, scantily dressed women, and other escapist content. While the Chinese government and online platforms are not 100 per cent effective in their ability to completely stamp out such content, the default is that such content is actively restricted, while US platforms tent to take a more pro-speech regulatory position [@liu2022]. If one of the main mechanisms through which social media promotes social isolation occurs is exposure to negative content, then this highly regulated online environment should lead to a weaker relationship between social media use and isolation. Suggestive of this is a a study by Liu and Liu that finds Chinese who consumed more commercialized and sensationalized media reported more trauma during the first COVID-19 outbreak in China than those who followed government sources [@liu2020].

On the other hand, China also has a highly robust online influencer culture. While social media influencers are a global phenomena, China has perhaps the most commercialized version of influencing. Social media influencers are developed by large agencies and work in an almost assembly-line fashion to promote products on livestreams or pitch sponsor products. In this influencer culture, less attention is paid to authenticity and personality and relatively more emphasis is placed on consumerism, brand promotion, and image [@tam2019; @wei2023]. Given that previous research has suggested one of the main channels through which social media creates feelings of isolation is negative personal comparisons with online content (so-called "FOMO" - fear of missing out), heavy social media use may increase the strength of the relationship between social media use and social isolation through consumption of such highly commercialized content.

@@

-   **H1b:** social media use should have a negative connection to social isolation due to the dominance of influencer culture in China

@@

The results of our models of social isolation are presented in @tbl-social-isolation-models. The additive estimates in model (1) provide mixed results regarding whether internet communication is a positive force in deterring social isolation. Not surprisingly, there is a negative relationship between the number of digital connections (human interaction) folks make and feelings of social isolation. On the other hand, though, both moving online relationships offline and general social media use are positively related to feelings of social isolation. The latter is consistent with much of the literature outlined above, but the former is not. Perhaps, this is context driven. The results also indicate that those living in more urban environments tend to feel less socially isolated.

@@

In terms of our original hypotheses, we find support for the idea that China's social media environment is not particularly unique with respect to generating feelings of social isolation (**H1b**). Whether or not this is due to the influencer culture or other factors, we cannot say with our data but it at least does suggest that greater content regulation is not likely to moderate the amount of isolation that social media produces. We also find, as has been found in other studies, that users who use the internet to create new connections are less likely to feel isolated (**H2**). However, this does not seem to extend to those that use the internet to facilitate offline meetups. Finally, we find that the impact of the internet and social media vary significantly according to the level of urbanicity (**H3**) - this matches with previous findings that suggest that there is significant heterogeneity in the impact of social media on social isloation. All of these findings are generally in line with previous research on the subject, suggesting that the mechanisms that generate feelings of social isolation are universal and not so sensitive to different social and governing contexts.

@@

Cyberoptimists heralded the internet as means by which people could connect, build social capital, and some suggested this could deter social isolationism. Our results provide little evidence of this optimistic view of the internet and its ability to improve the quality of people's lives. We do find that some are more likely to interact with family and friends and that this does deter feelings of social isolation, but on the other hand, even though many move online relationships offline and connect through social media, doing so does not deter feelings of social isolation. Instead, such digital actively is actually associated with increased feelings of social isolation.

We also found no evidence that the internet could serve as bridge to build connections for those living outside of cities. The optimistic view here would be that the internet could lower the costs of making connections for those who have less social opportunity, those living in less populated areas, and as a result, help them to feel less isolated. We, in fact, found the opposite. Internet communication had either less positive effect or more negative effect on feelings of social isolation for those living in less urban spaces.

Our contribution here is that we look at some well-established theoretical arguments in a context outside that of where most of the research centered on these type questions has examined. While our study is far from the last word on the subject, these results suggest that the mechanisms that link social media use and social isolation are relatively consistent across countries.

>>>>

---
title: "Lonely and online: an investigation of social media use and social isolation in China"
format:
  tandf-pdf:
    keep-tex: true 
date: '`r format(Sys.time(), "%B %d, %Y")`'
abstract: "Early research and popular discourse on the promise of the internet's impact on social isolation was optimistic. Since these early predictions, research has become less positive, usually pointing to the negative role of social media, withn more recent studies focused on heterogeneous effects. Using data from a nationwide survey of Chinese citizens to gain leverage on this disagreement in the literature, our research finds support for those that have focused on heterogeneous effects - we find that urbancity is a key moderating variable. Additionally, our results lend support to a FOMO mechanism while they are not suggestive of a mechanism by which negative content is the channel through which social media generates feelings of social isolations. Overall, our findings suggest that institutional context may not play an important role in the social media - social isolation relationship."
keywords: [Social Isolation, Social Media, Digital Communication, China]
execute: 
  echo: false
  warning: false
keep-tex: true
bibliography: references.bib
csl: taylor-and-francis-apa.csl
author:
  - name: Jason Gainous
    affiliations:
      - name: University of Sharjah
        department: Department of Mass Communication
        city: Sharjah
        country: United Arab Emirates
    orcid: 0000-0002-3104-2313
    email: jgainous@sharjah.ac.ae
  - name: Andrew W. MacDonald
    affiliations:
      - name: Duke Kunshan University
        department: Division of Social Sciences
        city: Kunshan
        country: China
    orcid: 0000-0002-4695-1560
    email: andrew.macdonald@dukekunshan.edu.cn
    attributes:
      corresponding: true
---

@@

Research is quite mixed when it comes to identifying the relationship between social media use and feelings of isolationism. Early research suggested that the internet would build social capital, or an interconnectedness between people, and that social media platforms were the ultimate way to bring people together. Since then, research on the subject has been decidedly more mixed. On one hand, some authors have found social media may encourage social isolation, and ultimately, social media users may feel more disconnected from others than those who do not use social media. On the other hand, others have found that there is only a minimal relationship between these two variables. More recent scholarship has suggested that social media may produce heterogeneous effects on individuals. This study uses the little-studied Chinese social media environment to try and gain leverage on this question - the Chinese online environment has certain features that should both moderate and exacerbate key causal pathways previously identified in the literature. To we employ a large survey of Chinese internet users and test a variety of regression specifications.[^1] The main result is that social media use is generally negatively related to social isolation, though the nature of this relationship changes depending on the urbanicity of the respondent - being in a rural area amplifies this relationship. This finding lends support both to the previous negative relationship findings and also reinforce the findings that emphasize heterogeneous effects.

@@

The existing literature on social isolation and social media use generally finds a negative to null relationship between the two variables, while the most recent research has suggested that social media has heterogeneous effects. Based on this existing literature, we then develop some testable hypotheses regarding how the relationship between usage and isolation may vary in the context of China. The result of these hypotheses tests will help to better understand the possible mechanism that links these two variables.

@@

The most important of these contextual factor in China is that social media space that is highly controlled online environment. Many people are aware that the Chinese government heavily censors politically sensitive content on online platforms. Fewer are aware that the Chinese government spends as much, if not more, effort to clean up non-political negative or distressing content on the Chinese internet, such as scams, instances of cyber bullying, disturbing videos, and vulgar content [@wang2020]. This control impulse also extends content the government views as spiritually polluting, such as pornography, scantily dressed women, and other escapist content. While the Chinese government and online platforms are not 100 per cent effective in their ability to completely stamp out such content, the default is that such content is actively restricted, while US platforms tent to take a more pro-speech regulatory position [@liu2022]. If one of the main mechanisms through which social media promotes social isolation occurs is exposure to negative content, then this highly regulated online environment should lead to a weaker relationship between social media use and isolation. Suggestive of this is a a study by Liu and Liu that finds Chinese who consumed more commercialized and sensationalized media reported more trauma during the first COVID-19 outbreak in China than those who followed government sources [@liu2020].

On the other hand, China also has a highly robust online influencer culture. While social media influencers are a global phenomena, China has perhaps the most commercialized version of influencing. Social media influencers are developed by large agencies and work in an almost assembly-line fashion to promote products on livestreams or pitch sponsored products. In this influencer culture, less attention is paid to authenticity and personality and relatively more emphasis is placed on consumerism, brand promotion, and image [@tam2019; @wei2023]. Given that previous research has suggested one of the main channels through which social media creates feelings of isolation is negative personal comparisons with online content (so-called "FOMO" - fear of missing out), heavy social media use may increase the strength of the relationship between social media use and social isolation through consumption of such highly commercialized content. These influencers are most popular among young consumers so one may expect that only younger users would experience these effects.

@@

-   **H1b:** social media use should have a strongly negative connection to social isolation due to the dominance of influencer culture in China, possibly moderated by age.

@@

The results of our models of social isolation are presented in @tbl-social-isolation-models. The additive estimates in model (1) provide mixed results regarding whether internet communication is a positive force in deterring social isolation. Not surprisingly, there is a negative relationship between the number of digital connections (human interaction) folks make and feelings of social isolation. On the other hand, though, both moving online relationships offline and general social media use are positively related to feelings of social isolation. The latter is consistent with much of the literature outlined above, but the former is not. The results also indicate that those living in more urban environments tend to feel less socially isolated.

@@

In terms of our original hypotheses, we find support for the idea that China's social media environment is not particularly unique with respect to generating feelings of social isolation (**H1b**). Whether or not this is due to the influencer culture or other factors, we cannot say with our data, however age is a consistently negative predictor of social isolation even when controlling for amount of social media use. This result is at least suggestive of the negative online influencer culture. We also find, as has been found in other studies, that users who use the internet to create new connections are less likely to feel isolated (**H2**). However, this does not seem to extend to those that use the internet to facilitate offline meetups. Finally, we find that the impact of the internet and social media vary significantly according to the level of urbanicity (**H3**) - this matches with previous findings that suggest that there is significant heterogeneity in the impact of social media on social isloation.

@@

Cyberoptimists heralded the internet as means by which people could connect, build social capital, and some suggested this could deter social isolationism. On the other hand, more recent research and popular media coverage of the subject has presented a deeply negative view of the role of social media. Our results do lean toward supporting the negative conclusions but add important contextualization and are suggestive of which of the hypothesized mechanisms are most relevant.

Firstly, we found no evidence that the internet could serve as bridge to build connections for those living outside of cities. The optimistic view that internet use could lower the costs of making connections for those who have less social opportunity, those living in less populated areas, and as a result, help them to feel less isolated. We, in fact, found the opposite. Internet communication had either less positive effect or more negative effect on feelings of social isolation for those living in less urban spaces. This finding lends support to the FOMO channel - viewing lots of content from influencers in a rural area compounds feelings of isolation.

Additionally, given the robustly negative relationship between social media use and social isolation even in the context of China's highly policed internet, our results cast some doubt on theories that suggest it is the negative content, such as cyberbullying or violent/graphic posts, that drives the relationship. Finally, our results also further suggest more work needs to be done in identifying important hetergeneous effects - in our results, those living in a big city experienced almost no change in social isolation as social media use increased while the relationship becomes increasingly negative by ruralness. This specific heterogeneous effect may be unique to China but the strength of it suggests that important inter-personal factors may be relevant in other contexts.

While our study is far from the last word on the subject, we hope this paper further helps move researchers away from considering the binary question of whether social media use has a negative relationship with social isolation to more productively considering the contextual and personal attributes that moderate and mediate this relationship.

### 2023-10-08 tandf/Social Isolation in China.qmd
prepared for t and f submission

---
title: "Lonely and online: an investigation of social media use and social isolation in China"
format:
  tandf-pdf:
    keep-tex: true 
date: '`r format(Sys.time(), "%B %d, %Y")`'
abstract: "Early research and popular discourse on the promise of the internet's impact on social isolation was optimistic. Since these early predictions, research has become less positive, usually pointing to the negative role of social media, withn more recent studies focused on heterogeneous effects. Using data from a nationwide survey of Chinese citizens to gain leverage on this disagreement in the literature, our research finds support for those that have focused on heterogeneous effects - we find that urbancity is a key moderating variable. Additionally, our results lend support to a FOMO mechanism while they are not suggestive of a mechanism by which negative content is the channel through which social media generates feelings of social isolations. Overall, our findings suggest that institutional context may not play an important role in the social media - social isolation relationship."
keywords: [Social Isolation, Social Media, Digital Communication, China]
execute: 
  echo: false
  warning: false
keep-tex: true
bibliography: references.bib
csl: taylor-and-francis-apa.csl
author:
  - name: Jason Gainous
    affiliations:
      - name: University of Sharjah
        department: Department of Mass Communication
        city: Sharjah
        country: United Arab Emirates
    orcid: 0000-0002-3104-2313
    email: jgainous@sharjah.ac.ae
  - name: Andrew W. MacDonald
    affiliations:
      - name: Duke Kunshan University
        department: Division of Social Sciences
        city: Kunshan
        country: China
    orcid: 0000-0002-4695-1560
    email: andrew.macdonald@dukekunshan.edu.cn
    attributes:
      corresponding: true
---

>>>>

---
title: "Lonely and online: an investigation of social media use and social isolation in China"
date: '`r format(Sys.time(), "%B %d, %Y")`'
abstract: "Early research and popular discourse on the promise of the internet's impact on social isolation was optimistic. Since these early predictions, research has become less positive, usually pointing to the negative role of social media, withn more recent studies focused on heterogeneous effects. Using data from a nationwide survey of Chinese citizens to gain leverage on this disagreement in the literature, our research finds support for those that have focused on heterogeneous effects - we find that urbancity is a key moderating variable. Additionally, our results lend support to a FOMO mechanism while they are not suggestive of a mechanism by which negative content is the channel through which social media generates feelings of social isolations. Overall, our findings suggest that institutional context may not play an important role in the social media - social isolation relationship."
keywords: [Social Isolation, Social Media, Digital Communication, China]
format:
  tandf-pdf:
    keep-tex: true 
execute: 
  echo: false
  warning: false
keep-tex: true
bibliography: references.bib
csl: taylor-and-francis-apa.csl
author:
  - name: Jason Gainous
    affiliations:
      - name: University of Sharjah
        department: Department of Mass Communication
        city: Sharjah
        country: United Arab Emirates
    orcid: 0000-0002-3104-2313
    email: jgainous@sharjah.ac.ae
  - name: Andrew W. MacDonald
    affiliations:
      - name: Duke Kunshan University
        department: Division of Social Sciences
        city: Kunshan
        country: China
    orcid: 0000-0002-4695-1560
    email: andrew.macdonald@dukekunshan.edu.cn
    attributes:
      corresponding: true
---
